Extreme heterogeneity of influenza virus transcription in single cells

### 2017-09-12 paper/paper.tex
some tweaks to Results

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Transcriptional dynamics of influenza virus infection in single cells}

@@

\subsection{Strategy to measure mRNA in single virus-infected cells.}
We performed single-cell mRNA sequencing using a droplet-based system that physically isolates individual cells prior to reverse transcription~\citep{zheng2017massively,Macosko:2015ht,Klein:2015ki}.
Each droplet is associated with a unique \emph{cell barcode} that tags all mRNAs from that cell during reverse-transcription.
A random \emph{unique molecular identifier (UMI)} is additionally appended to each mRNA molecule during reverse transcription.
The 3' ends of the mRNAs are sequenced and mapped to the human and influenza virus transcriptomes to determine transcript identities.
This information is combined with that provided by the UMIs and cell barcodes to quantify the number of molecules of each mRNA species that have been captured for each cell.

Infected cells will express viral as well as cellular mRNAs -- however the cell barcodes and UMIs cannot distinguish whether a cell was initially infected by one or multiple viral particles.
We therefore engineered an influenza virus (strain A/WSN/1933) that additionally carried \emph{viral barcodes} consisting of synonymous mutations near the 3' end of each transcript (Figure~\ref{fig:workflow}A).
Critically, these synonymous mutations did not greatly impact viral growth kinetics (Figure~\ref{fig:workflow}B).
We infected A549 human lung carcinoma cells with an equal mix of the wild-type and synonymously barcoded viruses.
Cells infected by a single virion will exclusively express mRNAs from either wild-type or synonymously barcoded virus, whereas cells that are co-infected with multiple virions will often express mRNAs from both the wild-type and synonymously barcoded viruses (Figure~\ref{fig:workflow}C).
\begin{figure}
\centerline{\includegraphics[width=0.8\linewidth]{figures/Workflow/workflow.pdf}}
\caption{\label{fig:workflow} Experimental design.
{\bf (A)}  
We engineered a virus that carried two synonymous mutations near the 3' end of each mRNA.
At top are the specific mutations for PB2.
At bottom are locations of the synonymous mutations relative to the typical distribution of read depth for our 3'-end sequencing.
{\bf (B)} 
The wild-type and synonymously barcoded viruses transcribe their genes with similar kinetics. 
Shown is the abundance of the viral hemagglutinin (HA) transcript relative to the cellular housekeeping gene L32 as assessed by qPCR in A549 cells infected at an MOI of 0.5 (as determined on MDCK-SIAT1 cells).
Error bars $\pm$ S.D., n=3.
{\bf (C)}  
A549 cells were infected with an equal mixture of mutant and wild-type virus. 
Immediately prior to collection, cells were physically separated into droplets and cDNA libraries were generated containing the indicated barcodes. 
The libraries were deep sequenced, and the data processed to create a matrix that gives the number of molecules of each transcript observed in each cell.
Infected cells were further annotated by whether their viral mRNAs derived from wild-type virus, synonymously barcoded virus, or both.
}
\figdata{Sequences of wild-type and barcoded viruses are in \texttt{viralsequences.fasta}.\label{figdata:viralsequences}}
\end{figure}

We took care to generate stocks of virus that were relatively ``pure'' of defective particles.
Stocks of viruses typically contain an array of biologically active viral particles, some of which are defective for replication owing to mutations or deletions in essential viral genes~\citep{von1954incomplete,huang1970defective,Brooke:2014ch,Fonville:2015cg,Lauring:2010if,Dimmock:2014dk}.
These defective particles become prevalent when a virus is grown at high multiplicity of infection (MOI), where complementation permits the growth of otherwise deleterious genotype.
To minimize the levels of defective particles, we propagated our viral stocks at low MOI for a relatively brief period of time~\citep{Xue:2016bw}.
We then validated that our stocks exhibited greater purity of infectious particles than a stock propagated at high MOI by verifying that they had a higher ratio of infectious particles to both virion RNA (Figure~\ref{fig:viruspopulations}A) and particles capable of inducing expression of a single viral protein (Figure~\ref{fig:viruspopulations}B).
In addition, viral stocks with many defective particles are more immunostimulatory~\citep{Tapia:2013kf,Lopez:2014en}.
We confirmed that our viral stocks induced much less interferon than a stock propagated at higher MOI (Figure~\ref{fig:viruspopulations}C).

@@

\subsection{Single cells show an extremely wide range of expression of viral mRNA.}
We infected A549 cells at low MOI with a mixture of the wild-type and synonymously barcoded viruses, and collected cells for sequencing at 6, 8, and 10 hours post-infection, including a replicate of the 8-hour timepoint.
We recovered between 3,000 and 4,000 cells for each sample (Figure~\ref{fig:cells}A). 
As expected for a low-MOI infection, most cells expressed little or no viral mRNA (Figure~\ref{fig:cells}B).
Also as expected, the amount of viral mRNA per cell among infected cells increased over time (Figure~\ref{fig:cells}B).
But what was most notable was how widely the number of viral mRNA molecules varied among infected cells.
While the fraction of mRNA derived from virus was $<$0.1\% for most cells, viral mRNA constituted close to half the transcriptome in a few cells at 8 and 10 hours (Figure~\ref{fig:cells}C).
%Therefore, there is extraordinary variability in the number of viral transcripts per infected cell.

@@

A complicating factor is that uninfected cells could have small amounts of viral mRNA due to leakage of transcripts from lysed cells.
It is therefore important to establish a threshold for identifying truly infected cells.
We can do this by taking advantage of the fact that roughly half the infecting virions bear synonymous barcodes.
Reads derived from lysed cells will be drawn from both wild-type and synonymously barcoded viral transcripts.
However, most cells are infected by at most one virion, and so the reads from truly infected cells will usually derive almost entirely from one of the two viral variants. 
Figure~\ref{fig:viralbarcodes}A shows the fraction of viral reads in individual cells from each viral variant, and Figure~\ref{fig:viralbarcodes}B indicates the fraction of viral reads from the most abundant variant in that cell.
Most cells with large amounts of viral mRNA have viral transcripts exclusively derived from one viral variant -- indicating non-random partitioning as expected from viral infection.
However, cells with a small amount of viral mRNA often have viral transcripts from both variants, as expected from the random partitioning associated with simple mRNA leakage.
Finally, a few cells with large amounts of viral mRNA have viral transcripts from both variants, reflecting co-infection with both variants.

We determined the threshold amount of viral mRNA per cell at which the barcode partitioning clearly resulted from infection rather than leakage (Figure~\ref{fig:viralbarcodes}C), and used this threshold to annotate cells that we were confident were truly infected.
We also annotated as co-infected cells above this threshold that had mRNA from both viral variants.
Figure~\ref{fig:viralbarcodes}D shows the number of cells annotated as infected and co-infected for each sample -- these cells are just a small fraction of the number of cells with any viral read.
These annotation thresholds are conservative, and will miss some true low-level infections, as well as any co-infections with the same viral variant.
However, it is important that the analyses below are restricted to cells that are truly infected with virus, so we accepted the loss of some low-level infections in order to avoid false positives.
Because the bulk of cells are not infected, we subsampled the uninfected cells to the numbers shown in Figure~\ref{fig:viralbarcodes}D to balance the proportions of infected and uninfected cells for all subsequent analyses.

Strikingly, the extreme variation in the number of viral transcripts per cell remains even after we apply these rigorous criteria for annotating infected cells (Figure~\ref{fig:viralbarcodes}E). 
The fraction of viral mRNA per infected cell follows a roughly exponential distribution, with many cells having few viral transcripts and a few cells having many.
In fact, a bare few infected cells contribute the majority of influenza transcripts.
Specifically, at 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $\approx$15\% of infected cells produce over half the viral transcripts (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu}).

@@

\subsection{Absence of viral genes partially explains cell-to-cell variability in viral transcript abundance.}
The influenza genome is segmented, and cells can fail to express a viral mRNA if the encoding gene segment is not packaged in the infecting virion or fails to initiate transcription after infection.
Indeed, several groups have reported that the majority of infected cells fail to express at least one viral gene~\citep{Brooke:2013kb, Heldt:2015iz,Dou:2017cp}. 
We wondered if the absence of specific viral genes might be associated with reduced amounts of viral mRNA within single infected cells.
In particular, transcription of influenza virus mRNAs is performed by the viral ribonucleoprotein (RNP) complex, which consists of the three proteins that encode the tripartite polymerase (PB2, PB1, and PA) as well as nucleoprotein (NP)~\cite{huang1990determination}.
Each viral gene segment is associated with one RNP in incoming infecting virions, but secondary transcription by newly synthesized RNPs requires the presence of the viral genes for each of the four constituent RNP proteins so that the cell can produce more RNP proteins~\citep{Vreede:2004ip,eisfeld2015centre}.
This secondary transcription is a major source of viral mRNAs, as evidenced by the fact that blocking synthesis of the RNP proteins reduces the amount of viral mRNA by several orders of magnitude in bulk cells (Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:cyclohexamide}).

\begin{figure}[t!]
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_burden_flu_gene_merge.pdf}}
\caption{\label{fig:fluburdenbyflugene}
The absence of viral genes explains some of the variability in the amount of viral mRNA per cell.
{\bf (A)} 
Fraction of mRNA in each infected cell derived from virus as a function of the normalized expression of each viral gene in that cell, taken over all time points.
Cells with high viral burden always express all the RNP genes, but some cells with high viral burden lack each of the other viral genes.
{\bf (B)}
Box plots showing the per-cell viral burden among cells with $>$0.5\% of their mRNA from virus, binned by whether or not the cells express each viral gene.
A Wilcoxon signed-rank test was used to test the null hypothesis that absence of each gene does not affect viral burden: **** = $P < 10^{-4}$, *** = $P < 10^{-3}$,  * = $P < 0.05$, ns = not significant.
Similar results are obtained if we examine only the 10-hour timepoint (Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:10hrfluburdenbyflugene}).
Top and bottom of box bound the first and third quartiles; the line in the box is at the median.
Whiskers extend to the highest or lowest point within1.5$\times$ the interquartile range, and outliers are shown as circles.
{\bf (C)}
Among cells that express all four RNP genes, this panel shows the fraction of that express each of the four other genes, as well as the fraction that express \emph{all} four of the other genes.
The fraction that express all four other genes is well predicted by simply multiplying the frequencies of cells that express each of these genes individually, indicating that gene absence is approximately independent across these four genes.
}
\figsupp[Secondary transcription from newly synthesized RNPs is a major source of viral mRNA during bulk infections.\label{figsupp:cyclohexamide}]{
A549 cells were infected at an MOI of 0.2 as calculated on MDCK-SIAT1 cels in either the presence or absence of the protein-translation inhibitor cyclohexamide, and viral mRNA was quantified at 8 hours post-infection by qPCR.
The cyclohexamide prevents translation of new PB2, PB1, PA, and NP protein, and so prevents the formation of the new RNPs needed for secondary transcription.
The bars show the relative amount of HA and PB2 mRNA in the absence versus the presence of cyclohexamide. Error $\pm$ S.D. n=3. 
}{\includegraphics[width=0.3\linewidth]{figures/Primary_transcription_measurements/Cyclohexamide.pdf}}
\figsupp[Like panel (B) but for the 10-hr sample only.\label{figsupp:10hrfluburdenbyflugene}]{
The absence of viral RNP genes but \emph{not} non-RNP genes remains significantly associated with reduced viral burden when we examine only the 10-hr sample, which is the single time point with the most data points.
The difference for NP is no longer statistically significant due to low counts of infected cells lacking NP, but the trend remains.
}{\includegraphics[width=\linewidth]{figures/p_10hr_flu_burden_flu_gene_test}}
\figdata{The numerical data for panel (C) are in \texttt{p\_missing\_genes.csv}.}
\end{figure}

@@

We also sought to quantify the fraction of infected cells that completely failed to express a given gene.
We limited this analysis to examining the presence / absence of the non-RNP genes in cells expressing all four RNP genes, since we might fail to detect viral transcripts that are actually present at low levels in RNP-deficient cells due to the lower viral burden in these cells.
At the 8- and 10-hour time points, between 5\% and 17\% of cells fail to express any one of the four non-RNP genes (Figure~\ref{fig:fluburdenbyflugene}C).
The absence of a given gene appears to be an independent event, as the probability of observing all four non-RNP genes in a cell is well predicted by simply multiplying the probabilities of observing each gene individually (Figure~\ref{fig:fluburdenbyflugene}C). 
Our measurements of the frequency at which infected cells fail to express individual viral genes are roughly consistent with estimates made by others using different means~\citep{Brooke:2013kb,Dou:2017cp}.

\subsection{The relative amounts of different viral mRNAs are fairly consistent across cells.}
The results above show that the total amount of viral mRNA in infected cells varies over orders of magnitude.
Does the relative expression of viral genes exhibit similar cell-to-cell variability?
To address this question, we focused on cells that derived $>$5\% of their mRNA from virus, since estimates of relative viral gene expression will be less noisy in cells with more viral mRNAs.

@@

In contrast to the extreme variability in the total viral mRNA per cell, the fraction of this mRNA derived from each gene is much more consistent across cells (Figure~\ref{fig:fluexpr}A).
Total viral mRNA varies by orders of magnitude, but the fraction from any given viral gene is fairly tightly clustered around the median value for all cells (Figure~\ref{fig:fluexpr}B).
The relative levels of each viral mRNA in our cells are similar to prior bulk measurements made by Northern blots~\citep{Hatada:1989vz}, which also found an expression hierarchy of M $>$ NS $\gg$ NP $>$ NA $>$ HA $\gg$ PB2 $\sim$ PB1 $\sim$ PA.
The cell-to-cell consistency in the relative expression of different viral genes generally becomes even tighter if we limit the analysis only to cells that express all eight viral genes (Figure~\ref{fig:fluexpr}C,D).
Therefore, with the exception of complete gene absence, the factors that drive the dramatic cell-to-cell variability in the amount of viral mRNA appear to have roughly similar effects on all viral genes in a given cell.
This finding is consistent with prior work showing positive correlations among the abundance of several viral gene segments in individual cells ~\citep{Heldt:2015iz}.

@@

\subsection{Activation of the interferon response is rare in single infected cells.}
Because our sequencing captured all polyadenylated transcripts, we can examine whether there are prominent changes in the host-cell transcriptome in sub-populations of infected cells.
In particular, influenza virus infection can trigger innate-immune sensors that lead to the transcriptional induction of type I and III interferons, and subsequently of anti-viral interferon-stimulated genes~\citep{Killip:2015dw,Iwasaki:2014dw,Crotta:2013ef}.
In bulk cells, this interferon response is one of the most prominent transcriptional signatures of influenza-virus infection~\citep{Geiss:2002a}.
However, activation of the interferon response is stochastic and bi-modal at the level of single cells~\citep{Chen:2010cr,Shalek:2014ey,Shalek:2013ej,PerezCidoncha:2014jr,bhushal2017cell}.
We therefore hypothesized that we might see two sub-populations of infected cells: one in which the interferon response inhibited viral transcription, and another in which the virus was able to express high levels of its mRNA by evading or blocking this response.

\begin{figure}
\centerline{\includegraphics[width=0.8\linewidth]{figures/p_small_tsne_merge.pdf}}
\caption{\label{fig:tsne}
A t-SNE plot created by semi-supervised clustering of the cells using genes that co-vary with viral infection status.
Each point is a single cell, and each panel shows an identical layout but colors the cells according to a different property.
{\bf (A), (B)}
Cells colored by the fraction of their mRNA that is derived from virus.
{\bf (C)}
Cells colored by the experimental sample from which they are derived.
{\bf (D), (E)}
Cells colored by the number of detected transcripts from type I and III interferons (IFN).
Only one cell has detectable interferon expression.
{\bf (E)}
Cells colored by the expression of the interferon-stimulated gene IFIT1.
{\bf (F)}
Cells colored by whether they express the viral NS gene.
The one interferon-positive cell is lacking NS, but many interferon-negative cells also lack NS.
}

@@

We posited that the paucity of interferon induction might be due to the activity of influenza virus's major interferon antagonist, the NS1 protein.
We therefore identified cells that expressed substantial amounts of viral mRNA but lacked the NS gene (Figure~\ref{fig:tsne}F).
Consistent with the idea that NS1 is important for suppressing interferon, the one interferon-positive cell lacked detectable expression of the NS gene.
But other cells that lacked NS expression still failed to induce a detectable interferon response, despite often having $>$10\% of their mRNA derived from virus (Figure~\ref{fig:tsne}).
Therefore, most infected cells fail to express interferon even when the virus lacks its major known interferon antagonist.
Our results are in line with other work showing that NS1-deficient influenza virus fails to deterministically induce interferon~\citep{Killip:2017ef,Kallfass:2013kp}.
Therefore, single cells infected at low MOI with a relatively ``pure'' stock of influenza virus clearly activate innate-immune responses much less frequently than might be presumed from bulk studies that flood cells with viral stocks full of defective particles.

\subsection{Genes involved in the oxidative-stress response co-vary with viral gene expression.}
We examined whether any host genes were differentially expressed in cells with more viral mRNA.
We restricted this analysis to infected cells with all eight viral genes in order to focus on cellular genes that were associated with viral mRNA burden independent of any effects due to the stochastic presence or absence of particular viral transcripts.
We identified 45 cellular genes that co-varied with viral gene expression at a false discovery rate of 0.1 (Figure~\ref{fig:cellulargenes}).

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Extreme heterogeneity of influenza virus transcription in single cells}

@@

\subsection{Strategy to measure mRNA in single virus-infected cells.}
We performed single-cell mRNA sequencing using a droplet-based system that physically isolates individual cells prior to reverse transcription~\citep{zheng2017massively,Macosko:2015ht,Klein:2015ki}.
Each droplet contains primers with a unique \emph{cell barcode} that tags all mRNAs from that droplet during reverse-transcription.
Each primer also contains a \emph{unique molecular identifier (UMI)} that is appended to each mRNA molecule during reverse transcription.
The 3' ends of the mRNAs are sequenced and mapped to the human and influenza virus transcriptomes to determine transcript identities.
This information is combined with that provided by the UMIs and cell barcodes to quantify the number of molecules of each mRNA species that have been captured for each cell.

Infected cells will express viral as well as cellular mRNAs -- however the cell barcodes and UMIs cannot distinguish whether a cell was initially infected by one or multiple viral particles.
We therefore engineered an influenza virus (strain A/WSN/1933) that additionally carried \emph{viral barcodes} consisting of synonymous mutations near the 3' end of each transcript (Figure~\ref{fig:workflow}A).
Critically, these synonymous mutations did not greatly impact viral growth kinetics (Figure~\ref{fig:workflow}B).
We infected A549 human lung carcinoma cells with an equal mix of the wild-type and synonymously barcoded viruses.
Cells infected by a single virion will exclusively express mRNAs from either wild-type or synonymously barcoded virus, whereas cells that are co-infected with multiple virions will often express mRNAs from both the wild-type and synonymously barcoded viruses (Figure~\ref{fig:workflow}C).
\begin{figure}
\centerline{\includegraphics[width=0.8\linewidth]{figures/Workflow/workflow.pdf}}
\caption{\label{fig:workflow} Experimental design.
{\bf (A)}  
We engineered a virus that carried two synonymous mutations near the 3' end of each mRNA.
At top are the specific mutations for PB2.
At bottom are locations of the synonymous mutations relative to the typical distribution of read depth for our 3'-end sequencing.
{\bf (B)} 
The wild-type and synonymously barcoded viruses transcribe their genes with similar kinetics. 
Shown is the abundance of the viral hemagglutinin (HA) transcript relative to the cellular housekeeping gene L32 as assessed by qPCR in A549 cells infected at an MOI of 0.5 (as determined on MDCK-SIAT1 cells).
Error bars $\pm$ S.D., n=3.
{\bf (C)}  
For the single-cell mRNA sequencing, A549 cells were infected with an equal mixture of mutant and wild-type virus. 
Immediately prior to collection, cells were physically separated into droplets and cDNA libraries were generated containing the indicated barcodes. 
The libraries were deep sequenced, and the data processed to create a matrix that gives the number of molecules of each transcript observed in each cell.
Infected cells were further annotated by whether their viral mRNAs derived from wild-type virus, synonymously barcoded virus, or both.
}
\figdata{Sequences of wild-type and barcoded viruses are in \texttt{viralsequences.fasta}.\label{figdata:viralsequences}}
\end{figure}

We took care to generate stocks of virus that were relatively ``pure'' of defective particles.
Stocks of viruses typically contain an array of biologically active viral particles, some of which are defective for replication owing to mutations or deletions in essential viral genes~\citep{von1954incomplete,huang1970defective,Brooke:2014ch,Fonville:2015cg,Lauring:2010if,Dimmock:2014dk}.
These defective particles become prevalent when a virus is grown at high multiplicity of infection (MOI), where complementation permits the growth of otherwise deleterious genotypes.
To minimize the levels of defective particles, we propagated our viral stocks at low MOI for a relatively brief period of time~\citep{Xue:2016bw}.
We validated that our stocks exhibited greater purity of infectious particles than a stock propagated at high MOI by verifying that they had a higher ratio of infectious particles to virion RNA (Figure~\ref{fig:viruspopulations}A) and to particles capable of inducing expression of a single viral protein (Figure~\ref{fig:viruspopulations}B).
In addition, viral stocks with many defective particles are more immunostimulatory~\citep{Tapia:2013kf,Lopez:2014en}.
We confirmed that our viral stocks induced much less interferon than a stock propagated at higher MOI (Figure~\ref{fig:viruspopulations}C).

@@

\subsection{Single cells show an extremely wide range of expression of viral mRNA.}
We infected A549 cells at low MOI with a mixture of the wild-type and synonymously barcoded viruses, and collected cells for sequencing at 6, 8, and 10 hours post-infection, including a replicate of the 8-hour timepoint.
For all samples except for the replicate 8-hour timepoint, we replaced the infection inoculum with fresh media after one hour, thereby ensuring that most infection was initiated during a narrow time window.
We recovered between 3,000 and 4,000 cells for each sample (Figure~\ref{fig:cells}A). 
As expected for a low-MOI infection, most cells expressed little or no viral mRNA (Figure~\ref{fig:cells}B).
Also as expected, the amount of viral mRNA per cell among infected cells increased over time (Figure~\ref{fig:cells}B).
But what was most notable was how widely the number of viral mRNA molecules varied among infected cells.
While the fraction of mRNA derived from virus was $<$0.1\% for most cells, viral mRNA constituted half the transcriptome in a few cells at 8 and 10 hours (Figure~\ref{fig:cells}C).

@@

A complicating factor is that uninfected cells could have small amounts of viral mRNA due to leakage of transcripts from lysed cells.
It is therefore important to establish a threshold for identifying truly infected cells.
We can do this by taking advantage of the fact that roughly half the infecting virions bear synonymous barcodes.
Reads derived from lysed cells will be drawn from both wild-type and synonymously barcoded viral transcripts.
However, most cells are infected by at most one virion, and so the reads from truly infected cells will usually derive almost entirely from one of the two viral variants. 
Figure~\ref{fig:viralbarcodes}A shows the fraction of viral reads in individual cells from each viral variant, and Figure~\ref{fig:viralbarcodes}B indicates the fraction of viral reads from the most abundant variant in that cell.
Most cells with large amounts of viral mRNA have viral transcripts exclusively derived from one viral variant -- indicating non-random partitioning as expected from viral infection.
However, cells with a small amount of viral mRNA often have viral transcripts from both variants, as expected from the random partitioning associated with simple mRNA leakage.
Finally, a few cells with large amounts of viral mRNA have viral transcripts from both variants, reflecting co-infection.

We determined the threshold amount of viral mRNA per cell at which the barcode partitioning clearly resulted from infection rather than leakage (Figure~\ref{fig:viralbarcodes}C), and used this threshold to annotate cells that we were confident were truly infected.
We also annotated as co-infected cells above this threshold that had mRNA from both viral variants.
Figure~\ref{fig:viralbarcodes}D shows the number of cells annotated as infected and co-infected for each sample -- these cells are just a small fraction of the number of cells with any viral read.
These annotation thresholds are conservative, and will miss some true low-level infections, as well as any co-infections with the same viral variant.
However, it is important that the analyses below are restricted to cells that are truly infected with virus, so we accepted the loss of some low-level infections in order to avoid false positives.
Because most cells are not infected, we subsampled the uninfected cells to the numbers shown in Figure~\ref{fig:viralbarcodes}D to balance the proportions of infected and uninfected cells for all subsequent analyses.

Strikingly, the extreme variation in the number of viral transcripts per cell remains even after we apply these rigorous criteria for annotating infected cells (Figure~\ref{fig:viralbarcodes}E). 
The fraction of viral mRNA per infected cell follows a roughly exponential distribution, with many cells having few viral transcripts and a few cells having many.
At 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $\approx$15\% of infected cells produce over half the viral transcripts (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu}).

@@

\subsection{Absence of viral genes partially explains cell-to-cell variability in viral transcript abundance.}
The influenza genome is segmented, and cells can fail to express a viral mRNA if the encoding gene segment is not packaged in the infecting virion or fails to initiate transcription after infection.
Indeed, several groups have reported that the majority of infected cells fail to express at least one viral gene~\citep{Brooke:2013kb, Heldt:2015iz,Dou:2017cp}. 
We wondered if the absence of specific viral genes might be associated with reduced amounts of viral mRNA within single infected cells.
In particular, transcription of influenza virus mRNAs is performed by the viral ribonucleoprotein (RNP) complex, which consists of the three proteins that encode the tripartite polymerase (PB2, PB1, and PA) as well as nucleoprotein (NP)~\citep{huang1990determination}.
Each viral gene segment is associated with one RNP in incoming infecting virions, but secondary transcription by newly synthesized RNPs requires the presence of the viral genes encoding each of the four RNP proteins~\citep{Vreede:2004ip,eisfeld2015centre}.
This secondary transcription is a major source of viral mRNAs, as evidenced by the fact that blocking synthesis of the RNP proteins reduces the amount of viral mRNA by several orders of magnitude in bulk cells (Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:cyclohexamide}).

\begin{figure}[t!]
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_burden_flu_gene_merge.pdf}}
\caption{\label{fig:fluburdenbyflugene}
The absence of viral genes explains some of the variability in the amount of viral mRNA per cell.
{\bf (A)} 
Fraction of mRNA in each infected cell derived from virus as a function of the normalized expression of each viral gene in that cell, taken over all time points.
Cells with high viral burden always express all RNP genes, but some cells with high viral burden lack each of the other genes.
{\bf (B)}
Box plots showing the per-cell viral burden among cells with $>$0.5\% of their mRNA from virus, binned by whether or not the cells express each gene.
A Wilcoxon signed-rank test was used to test the null hypothesis that absence of each gene does not affect viral burden: **** = $P < 10^{-4}$, *** = $P < 10^{-3}$,  * = $P < 0.05$, ns = not significant.
Similar results are obtained if we examine only the 10-hour timepoint (Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:10hrfluburdenbyflugene}).
Top and bottom of box bound the first and third quartiles; the line in the box is at the median.
Whiskers extend to the highest or lowest point within1.5$\times$ the interquartile range, and outliers are shown as circles.
{\bf (C)}
The fraction of that express each of the four other genes among cells that express all RNP genes, as well as the fraction that express \emph{all} four of the other genes.
The fraction that express all four genes is well predicted by simply multiplying the frequencies of cells that express each gene individually, indicating that gene absence is approximately independent across these genes.
}
\figsupp[Secondary transcription from newly synthesized RNPs is a major source of viral mRNA during bulk infections.\label{figsupp:cyclohexamide}]{
A549 cells were infected at an MOI of 0.2 as calculated on MDCK-SIAT1 cels in either the presence or absence of the protein-translation inhibitor cyclohexamide, and viral mRNA was quantified at 8 hours post-infection by qPCR.
The cyclohexamide prevents translation of new PB2, PB1, PA, and NP protein, and so prevents the formation of the new RNPs needed for secondary transcription.
The bars show the relative amount of HA and PB2 mRNA in the absence versus the presence of cyclohexamide. Error $\pm$ S.D. n=3. 
}{\includegraphics[width=0.3\linewidth]{figures/Primary_transcription_measurements/Cyclohexamide.pdf}}
\figsupp[Like panel (B) but for the 10-hr sample only.\label{figsupp:10hrfluburdenbyflugene}]{
The absence of viral RNP genes but \emph{not} non-RNP genes remains significantly associated with reduced viral burden when we examine only the 10-hr sample, which is the single time point with the most data points.
The difference for NP is no longer statistically significant due to low counts of infected cells lacking NP, but the trend remains.
}{\includegraphics[width=\linewidth]{figures/p_10hr_flu_burden_flu_gene_test}}
\figdata{The numerical data for panel (C) are in \texttt{p\_missing\_genes.csv}.}
\end{figure}

@@

We also quantified the fraction of infected cells that completely failed to express a given gene.
We limited this analysis to examining the presence / absence of the non-RNP genes in cells expressing all four RNP genes, since we might fail to detect viral transcripts that are actually present at low levels in RNP-deficient cells due to the lower viral burden in these cells.
At the 8- and 10-hour time points, between 5\% and 17\% of cells fail to express any one of the four non-RNP genes (Figure~\ref{fig:fluburdenbyflugene}C).
The absence of a given gene appears to be an independent event, as the probability of observing all four non-RNP genes in a cell is well predicted by simply multiplying the probabilities of observing each gene individually (Figure~\ref{fig:fluburdenbyflugene}C). 
If we extrapolate the frequencies at which cells lack non-RNP genes to the RNP genes, then we would predict that 40-50\% of infected cells express mRNAs from all eight genes.
This estimate of the frequency at which infected cells express mRNAs from all eight gene segments is slightly higher than previous estimates of 13\%~\citep{Brooke:2013kb} and 20\%~\citep{Dou:2017cp}.
At least one difference is that \citet{Brooke:2013kb} stained for proteins whereas we examined the expression of mRNAs -- it is likely that some cells contain mutated viral genes that fail to produce stable protein even when mRNA is expressed.

\subsection{The relative amounts of different viral mRNAs are more consistent across cells.}
The results above show that the amount of viral mRNA in infected cells varies over several orders of magnitude.
Does the relative expression of viral genes exhibit similar cell-to-cell variability?
To address this question, we focused on cells that derived $>$5\% of their mRNA from virus, since estimates of relative viral gene expression will be less noisy in cells with more viral mRNAs.

@@

In contrast to the extreme variability in the total viral mRNA per cell, the fraction of this mRNA derived from each gene is much more consistent across cells (Figure~\ref{fig:fluexpr}A).
Total viral mRNA varies by orders of magnitude, but the fraction from any given viral gene is fairly tightly clustered around the median value for all cells (Figure~\ref{fig:fluexpr}B).
The relative levels of each viral mRNA in our cells are similar to prior bulk measurements made by Northern blots~\citep{Hatada:1989vz}, which also found an expression hierarchy of M $>$ NS $\gg$ NP $>$ NA $>$ HA $\gg$ PB2 $\sim$ PB1 $\sim$ PA.
The cell-to-cell consistency in the relative expression of different viral genes is even tighter if we limit the analysis only to cells that express all eight viral genes (Figure~\ref{fig:fluexpr}C,D).
Therefore, with the exception of complete gene absence, the factors that drive the dramatic cell-to-cell variability in the amount of viral mRNA have roughly similar effects on all viral genes in a given cell.
This finding is consistent with prior work showing positive correlations among the abundance of several viral genes in individual cells~\citep{Heldt:2015iz}.

@@

\subsection{Activation of the interferon response is rare in single infected cells.}
Because our sequencing captured all polyadenylated transcripts, we can examine whether there are prominent changes in the host-cell transcriptome in sub-populations of infected cells.
Influenza virus infection can trigger innate-immune sensors that lead to the transcriptional induction of type I and III interferons, and subsequently of anti-viral interferon-stimulated genes~\citep{Killip:2015dw,Iwasaki:2014dw,Crotta:2013ef}.
In bulk cells, this interferon response is one of the most prominent transcriptional signatures of influenza-virus infection~\citep{Geiss:2002a}.
However, activation of the interferon response is stochastic and bi-modal at the level of single cells~\citep{Chen:2010cr,Shalek:2014ey,Shalek:2013ej,PerezCidoncha:2014jr,bhushal2017cell}.
We therefore hypothesized that we might see two sub-populations of infected cells: one in which the interferon response inhibited viral transcription, and another in which the virus was able to express high levels of its mRNA by evading or blocking this response.

\begin{figure}
\centerline{\includegraphics[width=0.8\linewidth]{figures/p_small_tsne_merge.pdf}}
\caption{\label{fig:tsne}
A t-SNE plot created by semi-supervised clustering using genes that co-vary with viral infection status.
Each point is a single cell, and each panel shows an identical layout but colors the cells according to a different property.
{\bf (A), (B)}
Cells colored by the fraction of their mRNA derived from virus.
{\bf (C)}
Cells colored by the experimental sample.
{\bf (D), (E)}
Cells colored by the number of detected transcripts from type I and III interferons (IFN).
Only one cell has detectable interferon expression.
{\bf (E)}
Cells colored by the expression of the interferon-stimulated gene IFIT1.
{\bf (F)}
Cells colored by whether they express the viral NS gene.
The one interferon-positive cell is lacking NS, but many interferon-negative cells also lack NS.
}

@@

We posited that the paucity of interferon induction might be due to the activity of influenza virus's major interferon antagonist, the NS1 protein~\citep{garcia1998influenza,hale2008multifunctional}.
We therefore identified cells that expressed substantial amounts of viral mRNA but lacked the NS gene (Figure~\ref{fig:tsne}F).
Consistent with the idea that NS1 is important for suppressing interferon, the one interferon-positive cell lacked detectable expression of the NS gene.
But other cells that lacked NS expression still failed to induce a detectable interferon response, despite often having $>$10\% of their mRNA derived from virus (Figure~\ref{fig:tsne}).
Therefore, most infected cells fail to express interferon even when the virus lacks its major interferon antagonist.
Our results are in line with other work showing that NS1-deficient influenza virus fails to deterministically induce interferon~\citep{Killip:2017ef,Kallfass:2013kp}.
Therefore, single cells infected at low MOI with a relatively ``pure'' stock of influenza virus only rarely activate innate-immune responses.

\subsection{Genes involved in the oxidative-stress response co-vary with viral gene expression.}
We examined whether any host genes were differentially expressed in cells with more viral mRNA.
We restricted this analysis to infected cells with all eight viral genes in order to focus on cellular genes that were associated with viral mRNA burden independent of effects due to the presence or absence of particular viral transcripts.
We identified 45 cellular genes that co-varied with viral gene expression at a false discovery rate of 0.1 (Figure~\ref{fig:cellulargenes}).
